A Plasma Torus Around a Young Low-Mass Star

### 2025-03-21 paper/ms.tex
update

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figures~\ref{fig:spec}
and~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination (in which Lyman-series
photons are trapped by high optical depth, so that recombinations
cascade through H$\alpha$ and other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with 
%densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$ and 
temperatures between 1,000-10,000\,K,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987,Draine2011}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$ suggest
$n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
We have assumed a uniform density clump, and so this yields a total
gas mass $M_{\rm gas}\approx 8\times10^{16}$\,g.

>>>>

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figures~\ref{fig:spec}
and~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination (in which Lyman-series
photons are trapped by high optical depth, so that recombinations
cascade through H$\alpha$ and other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with 
%densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$ and 
temperatures between 1,000-10,000\,K,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987,Draine2011}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming the emitting volume is a homogeneous sphere of radius $r$, we can write
\begin{align}
  n_{\rm H} &= 
  %\frac{1}{x}
  %\left(
  %  \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha} h \nu_{\rm H\alpha} V }
  %\right)^{1/2},
  %\\
  %n_{\rm H} &=
  5 \cdot 10^{11}\,{\rm cm}^{-3}
  \left(
    \frac{0.5}{x}
  \right)
  \left( 
    \frac{ L_{\rm H\alpha} }{ 10^{27}\,{\rm erg\,s^{-1}} }
    \frac{ 10^{-13}\,{\rm cm^{3}\,s^{-1}} }{ \alpha^{\rm eff}_{\rm H\alpha} }
  \right)^{1/2}
  \left(
    \frac{ 0.04\,R_\odot }{ r }
  \right)^{3/2}.
\end{align}
Given that we have assumed a uniform density clump, this suggests a total
gas mass at the stated nominal parameters of $M_{\rm gas}\approx 8\times10^{16}$\,g.
